---
title: "Random Forest Model"
author: "Anushree Shivarudrappa"
date: "June 27, 2016"
output: pdf_document
---

# 1. Pre-Processing
```{r, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)
library(dplyr)
library(scales)
library(RColorBrewer)
library(tidyr)
library(caTools)
library(rpart)
library(rpart.plot)
library(ROCR)
library(randomForest)
library(tree)
library(caret)
library(e1071)
```

# 2. Data Loading
```{r, echo=FALSE}
setwd("/Users/anushreeshivarudrappa/Desktop/Spring Board/SB Capstone DeathRecords DataSet")
```
```{r, message=FALSE, warning=FALSE,results='hide'}
Death_US <- fread("DeathRecords.csv", header = T)

```

# 3. Selecting dataset for model

```{r}
#Create a new subgroup for "Age" variable
Death_US$AgeGroup <- cut(Death_US$Age, c(0,19,25,39,60,110))
levels(Death_US$AgeGroup) <- c("Teenager", "Young_Adult", "Adult", "Middle_Age",
                               "Senior_Citizens")
# separates natural death
Death_US_natural <- Death_US[Death_US$MannerOfDeath == 7, ]

```

## Select required variables

```{r}
require(MASS)
require(dplyr)
natural_sub <- Death_US_natural %>% dplyr::select(Education2003Revision, Sex, Age, AgeGroup,
                      InfantAgeRecode22, 
                      PlaceOfDeathAndDecedentsStatus, MaritalStatus, InjuryAtWork,
                      MannerOfDeath, 
                      Autopsy, ActivityCode, PlaceOfInjury, Icd10Code,CauseRecode358,
                      CauseRecode113, InfantCauseRecode130,CauseRecode39,
                      NumberOfEntityAxisConditions,NumberOfRecordAxisConditions,Race)

```

## Converting Character variable into Integer variable
```{r, warning=FALSE}
natural_sub$Sex <- as.integer(as.factor(natural_sub$Sex)) 
natural_sub$MaritalStatus <- as.integer(as.factor(natural_sub$MaritalStatus))
natural_sub$InjuryAtWork <- as.integer(as.factor(natural_sub$InjuryAtWork))
natural_sub$Autopsy <- gsub("n", "N", natural_sub$Autopsy)
natural_sub$Autopsy <- as.integer(as.factor(natural_sub$Autopsy))  
natural_sub$Icd10Code <- as.integer(as.factor(natural_sub$Icd10Code))  
natural_sub$AgeGroup <- as.integer(as.factor(natural_sub$AgeGroup)) 


# check for NA values if there remove it
sapply(natural_sub, function(x) sum(is.na(x)))
na_index <- natural_sub[, !is.na(natural_sub$AgeGroup)]
natural_sub <- natural_sub[na_index]


# Convert all the variable into factor except CauseRecode39 & Age
natural_sub <- as.data.frame(natural_sub)
CauseRecode39 <- natural_sub$CauseRecode39
natural_sub$CauseRecode39 <- NULL
Age <- natural_sub$Age
natural_sub$Age <- NULL

natural_sub[sapply(natural_sub, is.integer)] <- lapply(natural_sub[sapply(natural_sub, is.integer)], 
                                       as.factor)
natural_sub <- cbind(natural_sub,Age,CauseRecode39)

```

As we analyzed, the input model variables are " Age + InfantAgeRecode22 + 
PlaceOfDeathAndDecedentsStatus + MaritalStatus + ActivityCode + PlaceOfInjury + 
NumberOfRecordAxisConditions"

```{r,warning=FALSE}
# Lets divide the CauseRecode39 into 3 groups wrt count of entries in each levels
table(factor(natural_sub$CauseRecode39))

# model data with CauseRecode39 having entries less than 10000
modeldata_less <- natural_sub$CauseRecode39 %in% 
                                c(1,2,3,5,25,29,32,33,34,35,38,40,41,42)
modeldata_less <- natural_sub[modeldata_less, ]

# model data with CauseRecode39 having entries less than 100000 and more than 10000
modeldata_mid <- natural_sub$CauseRecode39 %in% 
                               c(6,7,9,10,11,12,13,14,16,17,20,23,26,27,30,31,36,39)
modeldata_mid <- natural_sub[modeldata_mid, ]

# model data with CauseRecode39 having entries more than 100000
modeldata_more <- natural_sub$CauseRecode39 %in% c(8,15,21,22,24,28,37)
modeldata_more <- natural_sub[modeldata_more, ]                               
  
```

# Model for the data with CauseRecode39 having entries less than 10000
```{r}
# We will do a random 89:11 split in our data set (89% will be for training models, 
# 11% to evaluate them)
set.seed(111)
# randomly pick 70% of the number of observations 
index <- sample.split(modeldata_less$CauseRecode39, SplitRatio = 0.89)
# subset data to include only the elements in the index
train <- subset(modeldata_less, index==T)
nrow(train)
# subset data to include all but the elements in the index
test <- subset(modeldata_less, index==F)
nrow(test)
# take a copy of ICD10Code of test set and remove the variable from test set
Cause39 <- test$CauseRecode39
test$CauseRecode39 <- NULL


## Model
model_forest <- randomForest(as.factor(CauseRecode39) ~  Age + InfantAgeRecode22 + 
                     PlaceOfDeathAndDecedentsStatus + MaritalStatus + ActivityCode + 
                       PlaceOfInjury + NumberOfEntityAxisConditions , 
                     data = train,
                     nodesize = 25, ntree = 1501)


# Predict the test dataset using random forest model
predict_forest <- predict(model_forest, newdata = test)
# confusion matrix
conf_matrix <- table(predict_forest, Cause39) 
conf_matrix
# accuracy
sum(diag(conf_matrix)) / nrow(test) 
```

# Model for the data with CauseRecode39 having entries more than 10000 and less than 100000
```{r}
# We will do a random 89:11 split in our data set (89% will be for training models, 
# 11% to evaluate them)
set.seed(112)
# randomly pick 70% of the number of observations 
index <- sample.split(modeldata_mid$CauseRecode39, SplitRatio = 0.89)
# subset data to include only the elements in the index
train <- subset(modeldata_mid, index==T)
nrow(train)
# subset data to include all but the elements in the index
test <- subset(modeldata_mid, index==F)
nrow(test)
# take a copy of ICD10Code of test set and remove the variable from test set
Cause39 <- test$CauseRecode39
test$CauseRecode39 <- NULL


## Model
model_forest <- randomForest(as.factor(CauseRecode39) ~  Age + InfantAgeRecode22 + 
                     PlaceOfDeathAndDecedentsStatus + MaritalStatus + ActivityCode + 
                       PlaceOfInjury + NumberOfEntityAxisConditions , 
                     data = train,
                     nodesize = 25, ntree = 1501, na.action = na.omit)

# Predict the test dataset using random forest model
predict_forest <- predict(model_forest, newdata = test)
# confusion matrix
conf_matrix <- table(predict_forest, Cause39) 
conf_matrix
# accuracy
sum(diag(conf_matrix)) / nrow(test) 

```

```{r, echo=FALSE}
rm(Death_US)
rm(Death_US_natural)
rm(natural_sub)
rm(modeldata_less)
rm(modeldata_less)
rm(train)
rm(test)
rm(model_forest)
rm(conf_matrix)

```

# Model for the data with CauseRecode39 having entries More than 100000
```{r}
# We will do a random 89:11 split in our data set (89% will be for training models, 
# 11% to evaluate them)
set.seed(113)
# randomly pick 70% of the number of observations 
index <- sample.split(modeldata_more$CauseRecode39, SplitRatio = 0.89)
# subset data to include only the elements in the index
train <- subset(modeldata_more, index==T)
nrow(train)
# subset data to include all but the elements in the index
test <- subset(modeldata_more, index==F)
nrow(test)
# take a copy of ICD10Code of test set and remove the variable from test set
Cause39 <- test$CauseRecode39
test$CauseRecode39 <- NULL

## Model
model_forest <- randomForest(as.factor(CauseRecode39) ~  Age + InfantAgeRecode22 + 
                     PlaceOfDeathAndDecedentsStatus + MaritalStatus + ActivityCode + 
                       PlaceOfInjury + NumberOfEntityAxisConditions , 
                     data = train,
                     nodesize = 25, ntree = 1501, na.action = na.omit)

# Predict the test dataset using random forest model
predict_forest <- predict(model_forest, newdata = test)
# confusion matrix
conf_matrix <- table(predict_forest, Cause39) 
conf_matrix
# accuracy
sum(diag(conf_matrix)) / nrow(test) 
```

 